\documentclass{exam}
\usepackage{amssymb,amsmath,listings,parskip}
\begin{document}
	
    \section{6.1: Counting}
    
    \textbf{Food truck}
    
    1 dollar: cupcake
    
    1 type of cake, 1 type of icing always.
    
    Cake: \{chocolate, blueberry, red velvet, peach, strawberry, vanilla\} (6 types)
    
    Icing: \{mango, vanilla, cream cheese, coffee\} (4 types)
    
    Product rule: $6 \cdot 4 = 24$\\
    
    \textbf{Pizza}
    
    3 dollars: pizza
    
    1 topping, sauce, crust always.
    
    Crust: \{thin, deep dish\} (2 types)
    
    Sauce: \{tomato, pesto, alfredo, BBQ\} (4 types)
    
    Topping: \{pineapple, pepperoni, mushrooms\} (3 types)
    
    (Extended) product rule: $2 \cdot 4 \cdot 3 = 24$\\
    
    1 free cupcake $\oplus$ 1 free pizza
    
    How many choices? Sum rule: $24 + 24 = 48$\\
    
    \textbf{Interns}
    
    12 offices (empty and unique)
    
    2 interns (Zim and Gir, unique)
    
    How many ways to have office assignments?
    
    If they must share an office: 12 possibilities.
    
    If they must NOT share an office: $12 \cdot 11$ possibilities.
    
    If sharing doesn't matter: $12 + 12 \cdot 11$ (there is no way to share and not share at the same time)\\
    
    \textbf{Car tag}
    
    Format: 3 letters and 4 numbers (LLLDDDD).
    
    26 possible letters, 10 possible numbers.
    
    Possibilities: $26^3 \cdot 10^4$
    
    If we want to avoid one certain sequence of 3 starting characters (that might be obscene), we would have $26^3 \cdot 10^4 - 10^4$ possibilities (inclusion-exclusion technique). Note we are not taking into account the spelling itself.\\
    
    How many bit strings are of length 7? $2^7$\\
    
    \textbf{BASIC language variable names}
    
    Rules: 
    
    1. Starts with a letter (one of 26 letters), and is otherwise alphanumeric (not case-sensitive).
    
    2. Maximum length is 2.
    
    3. Cannot use a reserved word. There are 5 "collisions" (reserved words of length 2).
    
    $V = V_1 + V_2$
    
    $V_1 = 26$
    
    $V_2 = 26 \cdot (26+10)$
    
    so $V = 962$
    
    Removing the 5 reserved word collisions, $V_f = 957$\\
    
    \end{document}
