\documentclass{exam}
\usepackage{amssymb,amsmath,listings,parskip}
\begin{document}
	
    \section{5.1: Mathematical induction (continued)}
    
    \textbf{Example.} $P(n): 3|(n^3 - n) \forall n,\ n \in \mathbb Z^+$
    
    Proof by math induction.
    
    Base step, $P(1)$:
    
    Let's look at $P(1): 3|(1^3 - 1)$
    
    $3|(1-1)$ (simplify)
    
    $3|0$ (simplify)
    
    By definiton, 3 divides 0, since $3 \cdot 0$ yields 0. Therefore, $P(1)$ is true.\\
    
    Inductive step: prove $P(k) \rightarrow P(k+1), k \in \mathbb Z^+$.
    
    Let's look at $P(k+1)$.
    
    $P(k+1) : 3|[(k+1)^3 - (k+1)]$
    
    $3|(k^3+3k^2+3k+1)-(k+1)$ Expand
    
    $3|(k^3 + 3k^2 + 3k - k)$ Regroup
    
    $3|(k^3 - k + 3k^2 + 3k)$ Regroup
    
    $3|[(k^3-k)+3(k^2+k)]$ Factor 3
    
    Note that by our inductive hypothesis, $P(k) (k^3-k)$ is divisible by 3. We conclude that $3(k^2+k)$ is divisible by 3 since it is a multiple of 3, $k^2 + k \in \mathbb Z$.
    
    The sum of two values divisible by 3 yields a sum divisible by 3.
    
    I've shown $P(k) \rightarrow P(k+1)$.
    
    Conclusion: since the base step and inductive step are both true, $P(n)$ is true $\forall n \in \mathbb Z^+$, by the principle of mathematical induction.\\
    
    {\large \textbf{Tiling}}
    
    Problem. $P(n)$: let $n \in \mathbb Z^+$. Show that every 2 by 2 board with one square removed can be tiled by right triangles.
    
    Proof by mathematical induction.
    
    Base case. $P(1): $ 2 by 2 board can be tiled having any one square removed as shown below.
    
    %TODO: add illustration
    
    Base case is complete.
    
    Inductive step. $P(k) \rightarrow P(k+1), k \in \mathbb Z^+$
    
    Let's look at a $2^{k+1} \times 2^{k+1}$ board.
    
    %TODO: add illustration
    
    \newpage
    
    \section{5.2: Strong induction}
    
    Recall the "ladder" from the mathematical induction section.
    
    With strong induction, it is possible to use more than 1 previous result to reach a desired next result.
    
    Take care no to go outside of your domain when going back.
    
\end{document}